Gender and Mental Depression: Analyzing Census Data Through Statistical Methods

### 2024-10-25 paper/paper.qmd
Following issue has been fixed: "Quick suggestions"

---
title: "Sex, education, and mental health: Analyzing American census data through statistical methods, 2018 to 2021"
subtitle: "A pre- and post-pandemic study with IPUM's MEPs data"
author: 
  - Christina Nguyen
  - Marcus Barnes
  - Nibras Ar Rakib
thanks: "Code and data are available at: [https://github.com/Nguyen-Ar-Rakib-Barnes/downturns-household-insurance-and-poverty](https://github.com/Nguyen-Ar-Rakib-Barnes/downturns-household-insurance-and-poverty)."
date: today
date-format: long
abstract: "Mental health poses a significant global health threat. This threat existed before the pandemic, and recent studies have shown that it has become even more prevalent after the pandemic. Our study developed a noble method to identify severe psychological depression using survey data that can be used as an early alert system to identify nonspecific psychological distress."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

library(boot)
library(broom.mixed)
library(collapse)
library(dataverse)
library(gutenbergr)
library(janitor)
library(knitr)
library(marginaleffects)
library(modelsummary)
library(rstanarm)
library(tidybayes)
library(tidyverse)
library(here)
library(ggplot2)
filtered_df_2018 <- read.csv(here::here("data/02-analysis_data/model_data_2018.csv"))
filtered_df_2021 <- read.csv(here::here("data/02-analysis_data/model_data_2021.csv"))

labeled_df_reduced_2018 <- filtered_df_2018 |> mutate(
  outcome = if_else(outcome==1, "Yes", "No"),
  outcome = as_factor(outcome),
  sex = if_else(sex==1,"Male", "Female"),
  sex = as_factor(sex),
  marital_status = case_when(
    marstat==10~"Married",
    marstat==20~"Widowed",
    marstat==30~"Divorced",
    marstat==40~"Separated",
    marstat==50~"Never married"
  ),
  marital_status = factor(marital_status, levels=c("Married","Widowed","Divorced","Separated","Never married")),
  race= case_when(
    racea==100~"White",
    racea==200~"Black/African-American",
    racea==300~"Aleut, Alaskan Native, or American Indian",
    racea==310~"Alaskan Native or American Indian",
    racea==320~"Alaskan Native/Eskimo",
    racea==330~"Aleut",
    racea==340~"American Indian",
    racea==350~"American Indian or Alaskan Native and any other group",
    racea==400~"Asian or Pacific Islander",
    racea==410~"Asian",
    racea==411~"Chinese",
    racea==412~"Filipino",
    racea==413~"Korean",
    racea==414~"Vietnamese",
    racea==415~"Japanese",
    racea==416~"Asian Indian",
    racea==420~"Pacific Islander",
    racea==421~"Hawaiian",
    racea==422~"Samoan",
    racea==423~"Guamanian",
    racea==430~"Other Asian or Pacific Islander",
    racea==431~"Other Asian or Pacific Islander (1992-1995)",
    racea==432~"Other Asian or Pacific Islander (1996)",
    racea==433~"Other Asian or Pacific Islander (1997-1998)",
    racea==434~"Other Asian (1999 forward)",
    racea==500~"Other Race",
    racea==510~"Other Race (1963-1977)",
    racea==520~"Other Race (1978)",
    racea==530~"Other Race (1979-1991)",
    racea==540~"Other Race (1992-1995)",
    racea==550~"Other Race (1996)",
    racea==560~"Other Race (1997-1998)",
    racea==570~"Other Race (1999-2002)",
    racea==580~"Primary Race not releasable",
    racea==600~"Multiple Race, No Primary Race Selected",
    racea==610~"Multiple Race, including Asian, excluding Black and White",
    racea==611~"Multiple Race, including Asian and Black, excluding White",
    racea==612~"Multiple Race, including Asian and White, excluding Black",
    racea==613~"Multiple Race, including Black, excluding Asian and White",
    racea==614~"Multiple Race, including Black and White, excluding Asian",
    racea==615~"Multiple Race, including White, excluding Asian and Black",
    racea==616~"Multiple Race, including Asian, White, and Black",
    racea==617~"Multiple Race, excluding Asian, White, and Black"
  ),
  race = factor(race, 
                levels=c("White", "Black/African-American", "Aleut, Alaskan Native, or American Indian", "Alaskan Native or American Indian", "Alaskan Native/Eskimo", "Aleut", "American Indian", "American Indian or Alaskan Native and any other group", "Asian or Pacific Islander", "Asian", "Chinese", "Filipino", "Korean", "Vietnamese", "Japanese", "Asian Indian", "Pacific Islander", "Hawaiian", "Samoan", "Guamanian", "Other Asian or Pacific Islander", "Other Asian or Pacific Islander (1992-1995)", "Other Asian or Pacific Islander (1996)", "Other Asian or Pacific Islander (1997-1998)", "Other Asian (1999 forward)", "Other Race", "Other Race (1963-1977)", "Other Race (1978)", "Other Race (1979-1991)", "Other Race (1992-1995)", "Other Race (1996)", "Other Race (1997-1998)", "Other Race (1999-2002)", "Primary Race not releasable", "Multiple Race, No Primary Race Selected", "Multiple Race, including Asian, excluding Black and White", "Multiple Race, including Asian and Black, excluding White", "Multiple Race, including Asian and White, excluding Black", "Multiple Race, including Black, excluding Asian and White", "Multiple Race, including Black and White, excluding Asian", "Multiple Race, including White, excluding Asian and Black", "Multiple Race, including Asian, White, and Black", "Multiple Race, excluding Asian, White, and Black"
                )),
  education = case_when(
    educ==100~"Grade 12 or less",
    educ==101~"Grade 12 or less",
    educ==102~"Never attended/kindergarten only",
    educ==103~"Grade 12 or less",
    educ==104~"Grade 12 or less",
    educ==105~"Grade 12 or less",
    educ==106~"Grade 12 or less",
    educ==107~"Grade 12 or less",
    educ==108~"Grade 12 or less",
    educ==109~"Grade 12 or less",
    educ==110~"Grade 12 or less",
    educ==111~"Grade 12 or less",
    educ==112~"Grade 12 or less",
    educ==113~"Grade 12 or less",
    educ==114~"Grade 12 or less",
    educ==115~"Grade 12 or less",
    educ==116~"Grade 12 or less",
    educ==200~"High school",
    educ==201~"High school",
    educ==202~"GED or equivalent",
    educ==300~"Some college",
    educ==301~"Some college",
    educ==302~"AA degree",
    educ==303~"AA degree",
    educ==400~"Bachelor's",
    educ==501~"Master's",
    educ==502~"Professional",
    educ==503~"Doctoral"
  ),
  education = factor(
    education,
    levels=c(
      "Grade 12 or less","Never attended/kindergarten only","High school","GED or equivalent","Some college","AA degree","Bachelor's","Master's","Professional","Doctoral"
    )
  )
)

labeled_df_reduced_2021 <- filtered_df_2021 |> mutate(
  outcome = if_else(outcome==1, "Yes", "No"),
  outcome = as_factor(outcome),
  sex = if_else(sex==1,"Male", "Female"),
  sex = as_factor(sex),
  marital_status = case_when(
    marstat==10~"Married",
    marstat==20~"Widowed",
    marstat==30~"Divorced",
    marstat==40~"Separated",
    marstat==50~"Never married"
  ),
  marital_status = factor(marital_status, levels=c("Married","Widowed","Divorced","Separated","Never married")),
  race= case_when(
    racea==100~"White",
    racea==200~"Black/African-American",
    racea==300~"Aleut, Alaskan Native, or American Indian",
    racea==310~"Alaskan Native or American Indian",
    racea==320~"Alaskan Native/Eskimo",
    racea==330~"Aleut",
    racea==340~"American Indian",
    racea==350~"American Indian or Alaskan Native and any other group",
    racea==400~"Asian or Pacific Islander",
    racea==410~"Asian",
    racea==411~"Chinese",
    racea==412~"Filipino",
    racea==413~"Korean",
    racea==414~"Vietnamese",
    racea==415~"Japanese",
    racea==416~"Asian Indian",
    racea==420~"Pacific Islander",
    racea==421~"Hawaiian",
    racea==422~"Samoan",
    racea==423~"Guamanian",
    racea==430~"Other Asian or Pacific Islander",
    racea==431~"Other Asian or Pacific Islander (1992-1995)",
    racea==432~"Other Asian or Pacific Islander (1996)",
    racea==433~"Other Asian or Pacific Islander (1997-1998)",
    racea==434~"Other Asian (1999 forward)",
    racea==500~"Other Race",
    racea==510~"Other Race (1963-1977)",
    racea==520~"Other Race (1978)",
    racea==530~"Other Race (1979-1991)",
    racea==540~"Other Race (1992-1995)",
    racea==550~"Other Race (1996)",
    racea==560~"Other Race (1997-1998)",
    racea==570~"Other Race (1999-2002)",
    racea==580~"Primary Race not releasable",
    racea==600~"Multiple Race, No Primary Race Selected",
    racea==610~"Multiple Race, including Asian, excluding Black and White",
    racea==611~"Multiple Race, including Asian and Black, excluding White",
    racea==612~"Multiple Race, including Asian and White, excluding Black",
    racea==613~"Multiple Race, including Black, excluding Asian and White",
    racea==614~"Multiple Race, including Black and White, excluding Asian",
    racea==615~"Multiple Race, including White, excluding Asian and Black",
    racea==616~"Multiple Race, including Asian, White, and Black",
    racea==617~"Multiple Race, excluding Asian, White, and Black"
  ),
  race = factor(race, 
                levels=c("White", "Black/African-American", "Aleut, Alaskan Native, or American Indian", "Alaskan Native or American Indian", "Alaskan Native/Eskimo", "Aleut", "American Indian", "American Indian or Alaskan Native and any other group", "Asian or Pacific Islander", "Asian", "Chinese", "Filipino", "Korean", "Vietnamese", "Japanese", "Asian Indian", "Pacific Islander", "Hawaiian", "Samoan", "Guamanian", "Other Asian or Pacific Islander", "Other Asian or Pacific Islander (1992-1995)", "Other Asian or Pacific Islander (1996)", "Other Asian or Pacific Islander (1997-1998)", "Other Asian (1999 forward)", "Other Race", "Other Race (1963-1977)", "Other Race (1978)", "Other Race (1979-1991)", "Other Race (1992-1995)", "Other Race (1996)", "Other Race (1997-1998)", "Other Race (1999-2002)", "Primary Race not releasable", "Multiple Race, No Primary Race Selected", "Multiple Race, including Asian, excluding Black and White", "Multiple Race, including Asian and Black, excluding White", "Multiple Race, including Asian and White, excluding Black", "Multiple Race, including Black, excluding Asian and White", "Multiple Race, including Black and White, excluding Asian", "Multiple Race, including White, excluding Asian and Black", "Multiple Race, including Asian, White, and Black", "Multiple Race, excluding Asian, White, and Black"
                )),
  education = case_when(
    educ==100~"Grade 12 or less",
    educ==101~"Grade 12 or less",
    educ==102~"Never attended/kindergarten only",
    educ==103~"Grade 12 or less",
    educ==104~"Grade 12 or less",
    educ==105~"Grade 12 or less",
    educ==106~"Grade 12 or less",
    educ==107~"Grade 12 or less",
    educ==108~"Grade 12 or less",
    educ==109~"Grade 12 or less",
    educ==110~"Grade 12 or less",
    educ==111~"Grade 12 or less",
    educ==112~"Grade 12 or less",
    educ==113~"Grade 12 or less",
    educ==114~"Grade 12 or less",
    educ==115~"Grade 12 or less",
    educ==116~"Grade 12 or less",
    educ==200~"High school",
    educ==201~"High school",
    educ==202~"GED or equivalent",
    educ==300~"Some college",
    educ==301~"Some college",
    educ==302~"AA degree",
    educ==303~"AA degree",
    educ==400~"Bachelor's",
    educ==501~"Master's",
    educ==502~"Professional",
    educ==503~"Doctoral"
  ),
  education = factor(
    education,
    levels=c(
      "Grade 12 or less","Never attended/kindergarten only","High school","GED or equivalent","Some college","AA degree","Bachelor's","Master's","Professional","Doctoral"
    )
  )
)

@@

This study therefore employs a statistical approach in R to examine the relationships between sex, educational attainment, income and reported cases of mental depression. An understanding of these dynamics contribute to a deeper understanding of how education and sex intersect to influence mental health, providing valuable insights for policymakers and mental health practitioners.

The key questions underpinning this research are: does higher education correlate with lower rates of depression? Are women disproportionately affected by depression compared to men at the same educational level? Can we develop a model that predicts the likelihood of psychological distress [@kessler_kessler_2012] given certain characteristics of an individual such as sex, education or age? To address these questions, we analyze national census data through various statistical methods, focusing on the intersection of gender, education, and mental health outcomes [@blewett_ipums_2023].

Predictive models - even simplified ones like the one we present below - are useful for understanding complex issues like mental health. They allow researchers and policymakers to quickly spot trends and make better decisions. By looking at how factors like sex and education connect to depression, these models show patterns which lets us identify at-risk groups early, and lets us provide tailored support and resources to those who need it most. Plus, predictive models help direct future research by pointing out where more exploration is needed, ensuring we keep learning and improving our approach to mental health. Being able to predict and tackle what affects mental health is key to building healthier communities.

@@

Our goal was to identify respondents, based on gender and education, who were suffering from non-specific psychological distress based on the Kessler scale before and after the COVID-19 pandemic. To support our analyses, we split into two datasets for 2018 and 2021. We also excluded the respondents whose ages are below 19 and over 80. After enforcing our inclusion criteria, we had 15716 respondents for 2018 and 12685 for 2021. To give a preview of our data, @tbl-2018 and @tbl-2021 show the first five rows for each row for 2018 and 2021, respectively.
```{r}
#| label: fig-education-number-rows
#| fig-cap: Education
#| warning: false
#| message: false
#| echo: false
#| results: hide
print(nrow(filtered_df_2018))
print(nrow(filtered_df_2021))
```
```{r}
#| label: tbl-2018
#| tbl-cap: "Year 2018"
#| warning: false
#| message: false
#| echo: false
kable(filtered_df_2018[1:5,c("year","age","sex","marstat","racea","educ","inctot","k6sum")])
```
```{r}
#| label: tbl-2021
#| tbl-cap: "Year 2021"
#| warning: false
#| message: false
#| echo: false
kable(filtered_df_2021[1:5,c("year","age","sex","marstat","racea","educ","inctot","k6sum")])
```
## Measurement

@@

The MEPS survey collects participants' K6 score for nonspecific psychological distress. The K6 score is a summed scale value, also known as the Kessler 6 Scale, which considers six manifestations of nonspecific psychological distress. A score of 13 or higher indicates likely severe mental illness [@kessler_kessler_2012]. The variable called K6SUM represents the score in our dataset [-@sec-appendix-data-details]. We constructed a new variable called "outcome," where 0 represents the score below 13 and 1 represents the score above 13. We considered zero as a person who is not experiencing psychological distress, denoted as "No," and one as a person experiencing severe psychological distress, denoted as "Yes." @fig-respondent-outcome represents the number of respondents we have in the dataset who were suffering from psychological distress. In 2018, 601 respondents were suffering severe psychological distress, where the number of total respondents was 15716. In 2021, 573 respondents were suffering from nonspecific psychological distress over the 30-day recall period, where the total number of respondents was 12685.

```{r}
#| label: fig-respondent-outcome
#| fig-cap: Number of Respondents Suffering Severe Psychological Distress
#| fig-subcap: 
#|  - "2018"
#|  - "2021"
#| layout-ncol: 2
#| echo: false
#| warning: false
max_count <- max(table(labeled_df_reduced_2018$outcome)) + 1000
labeled_df_reduced_2018 |> 
  ggplot(aes(x = outcome)) +
  geom_histogram(stat = "count", fill = "#013065", color = "white") +
  scale_y_continuous(limits = c(0, max_count)) +
  geom_text(
    stat = "count", 
    color = "black",
    aes(label = after_stat(count)), 
    vjust = -0.6,  # Adjust vertical position of labels
    size = 5       # Adjust size of labels
  ) +
  labs(title = "", 
       x = "Outcome", 
       y = "Frequency") +
  theme_light()+
  theme(
    plot.title = element_text(size = 20, face = "bold"),  # Title font size
    axis.title.x = element_text(size = 16),               # X-axis label size
    axis.title.y = element_text(size = 16),               # Y-axis label size
    axis.text = element_text(size = 14)                   # Axis text size (ticks)
  )

max_count <- max(table(labeled_df_reduced_2021$outcome)) + 1000
labeled_df_reduced_2021 |> 
  ggplot(aes(x = outcome)) +
  geom_histogram(stat = "count", fill = "#013065", color = "white") +
  scale_y_continuous(limits = c(0, max_count)) +
  geom_text(
    stat = "count", 
    color = "black",
    aes(label = after_stat(count)), 
    vjust = -0.3,  # Adjust vertical position of labels
    size = 5       # Adjust size of labels
  ) +
  labs(title = "", 
       x = "Outcome", 
       y = "Frequency") +
  theme_light()+
  theme(
    plot.title = element_text(size = 20, face = "bold"),  # Title font size
    axis.title.x = element_text(size = 16),               # X-axis label size
    axis.title.y = element_text(size = 16),               # Y-axis label size
    axis.text = element_text(size = 14)                   # Axis text size (ticks)
  )
```

@@

Our dataset has 8209 female respondents and 7507 male respondents for 2018. In 2021, the total number of respondents is lower compared to 2018. We have 6784 female respondents and 5901 male respondents for 2021 [@fig-sex].

```{r}
#| label: fig-sex
#| fig-cap: Distribution of Respondents by Sex
#| fig-subcap: 
#|  - "2018"
#|  - "2021"
#| layout-ncol: 2
#| echo: false
#| warning: false

max_count <- max(table(labeled_df_reduced_2018$sex)) + 1000
labeled_df_reduced_2018 |> 
  ggplot(aes(x = sex)) +
  geom_histogram(stat = "count", fill = "#013065", color = "white") +
  scale_y_continuous(limits = c(0, max_count)) +
  geom_text(
    stat = "count", 
    color = "black",
    aes(label = after_stat(count)), 
    vjust = -0.3,  # Adjust vertical position of labels
    size = 5       # Adjust size of labels
  ) +
  labs(title = " ", 
       x = "Sex", 
       y = "Frequency") +
  theme_light()+
  theme(
    plot.title = element_text(size = 20, face = "bold"),  # Title font size
    axis.title.x = element_text(size = 16),               # X-axis label size
    axis.title.y = element_text(size = 16),               # Y-axis label size
    axis.text = element_text(size = 14)                   # Axis text size (ticks)
  )
max_count <- max(table(labeled_df_reduced_2021$sex)) + 1000
labeled_df_reduced_2021 |> 
  ggplot(aes(x = sex)) +
  geom_histogram(stat = "count", fill = "#013065", color = "white") +
  scale_y_continuous(limits = c(0, max_count)) +
  geom_text(
    stat = "count", 
    color = "black",
    aes(label = after_stat(count)), 
    vjust = -0.3,  # Adjust vertical position of labels
    size = 5       # Adjust size of labels
  ) +
  labs(title = " ", 
       x = "Sex", 
       y = "Frequency") +
  theme_light()+
  theme(
    plot.title = element_text(size = 20, face = "bold"),  # Title font size
    axis.title.x = element_text(size = 16),               # X-axis label size
    axis.title.y = element_text(size = 16),               # Y-axis label size
    axis.text = element_text(size = 14)                   # Axis text size (ticks)
  )
```

@@

@fig-education-2018 and @fig-education-2021 depict the overall education status of our respondents. For both 2018 and 2022, a majority of respondents have at least a high school degree, followed by a bachelor's degree.
```{r}
#| label: fig-education-2018
#| fig-cap: Distribution of Respondents by Education (2018)
#| echo: false
#| warning: false
max_count <- max(table(labeled_df_reduced_2018$education)) + 1000
labeled_df_reduced_2018 |> 
  ggplot(aes(x = education)) +
  geom_histogram(stat = "count", binwidth = 5, fill = "#013065", color = "white") +
  scale_y_continuous(limits = c(0, max_count)) +
  geom_text(
    stat = "count", 
    color = "black",
    aes(label = after_stat(count)), 
    vjust = -0.3,  # Adjust vertical position of labels
    size = 3       # Adjust size of labels
  ) +
  labs(title = "Education", 
       x = "Education", 
       y = "Frequency") +
  theme_light()+
  theme(
    plot.title = element_text(size = 15, face = "bold"),  # Title font size
    axis.title.x = element_text(size = 10),               # X-axis label size
    axis.title.y = element_text(size = 10),               # Y-axis label size
    axis.text = element_text(size = 10),              # Axis text size (ticks)
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

@@

max_count <- max(table(labeled_df_reduced_2021$education)) + 1000
labeled_df_reduced_2021 |> 
  ggplot(aes(x = education)) +
  geom_histogram(stat = "count", binwidth = 5, fill = "#013065", color = "white") +
  scale_y_continuous(limits = c(0, max_count)) +
  geom_text(
    stat = "count", 
    color = "black",
    aes(label = after_stat(count)), 
    vjust = -0.3,  # Adjust vertical position of labels
    size = 3       # Adjust size of labels
  ) +
  labs(title = "Education", 
       x = "Education", 
       y = "Frequency") +
  theme_light()+
  theme(
    plot.title = element_text(size = 15, face = "bold"),  # Title font size
    axis.title.x = element_text(size = 10),               # X-axis label size
    axis.title.y = element_text(size = 10),               # Y-axis label size
    axis.text = element_text(size = 10),              # Axis text size (ticks)
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```
\newpage

@@

```{r}
#| label: fig-age
#| fig-cap: Distribution of Respondents by Education (2021)
#| echo: false
#| warning: false

max_count <- max(table(labeled_df_reduced_2021$age)) + 1000
labeled_df_reduced_2021 |> 
  ggplot(aes(x = age)) +
  geom_histogram(fill = "#013065", color = "white") +
  scale_y_continuous(limits = c(0, max_count)) +
  labs(title = "Age", 
       x = "Age", 
       y = "Frequency") +
  theme_light()+
  theme(
    plot.title = element_text(size = 15, face = "bold"),  # Title font size
    axis.title.x = element_text(size = 10),               # X-axis label size
    axis.title.y = element_text(size = 10),               # Y-axis label size
    axis.text = element_text(size = 10),              # Axis text size (ticks)
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```
\newpage

@@

We leverage the logistic regression model to identify the association between suffering from psychological distress with sex and education. We set our model following the @tellingstories guidelines with the rstanarm package [@rstanarm_goodrich, @rstanarm_brilleman] in R [@citer].

@@

We have not found any notable differences between education and psychological distress. Research indicates a strong connection between emotional well-being and education, as emotional learning is believed to be nurtured within the school environment and through the curriculum [@norwich_critical_2022;@billington_conditions_2022].

@@

We randomly selected 5000 samples for our model. We have not used any prediction to find out how good our model is for detecting psychological depression. We also have not conducted any data augmentation to improve our model result. Our next steps are carefully curating the dataset, introducing several data validation techniques, and increasing the sample set.

>>>>

---
title: "Gender and Mental Depression: Analyzing Census Data Through Statistical Methods"
author: 
  - Christina Nguyen
  - Marcus Barnes
  - Nibras Ar Rakib
thanks: "Code and data are available at: [https://github.com/Nguyen-Ar-Rakib-Barnes/downturns-household-insurance-and-poverty](https://github.com/Nguyen-Ar-Rakib-Barnes/downturns-household-insurance-and-poverty)."
date: today
date-format: long
abstract: "Mental health poses a significant global health threat. This threat existed before the pandemic, and recent studies have shown that it has become even more prevalent after the pandemic. Our study developed a method to identify severe psychological depression using survey data that can be used as an early alert system to identify nonspecific psychological distress."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

library(modelsummary)
library(rstanarm)
library(tidyverse)
library(here)
library(arrow)
library(knitr)

filtered_df_2018 <- read_parquet(here::here("data/02-analysis_data/model_data_2018.parquet"))
filtered_df_2021 <- read_parquet(here::here("data/02-analysis_data/model_data_2021.parquet"))

labeled_df_reduced_2018 <- filtered_df_2018

labeled_df_reduced_2021 <- filtered_df_2021
```

@@

This study employs a statistical approach in R to examine the relationships between gender, educational attainment, income and reported cases of mental depression. By exploring these dynamics, we hope to contribute to a deeper understanding of how education and gender intersect to influence mental health, providing valuable insights for policymakers and mental health practitioners.

Therefore our key questions underpinning this research include: does higher education correlate with lower rates of depression? Are women disproportionately affected by depression compared to men at the same educational level? Can we develop a model that predicts the likelihood of psychological distress [@kessler_kessler_2012] given certain characteristics of an individual such as sex, education or age? To address these questions, we analyze national census data through various statistical methods, focusing on the intersection of gender, education, and mental health outcomes [@blewett_ipums_2023].

Predictive models - even simplified ones like the one we present below - are helpful for understanding complex issues like mental health. They allow researchers and policymakers spot trends and make smarter decisions. By looking at how factors like gender and education connect to depression, these models show patterns which lets us identify at-risk groups early, and lets us provide tailored support and resources to those who need it most. Plus, predictive models help direct future research by pointing out where more exploration is needed, ensuring we keep learning and improving our approach to mental health. In a time when mental well-being is more important than ever, being able to predict and tackle what affects it is key to building healthier communities.

@@

Our goal was to identify respondents, based on gender and education, who were suffering from non-specific psychological distress based on the Kessler scale before and after the COVID-19 pandemic. To support our analyses, we split into two datasets for 2018 and 2021. We excluded the respondents whose ages are below 19 and over 80. After enforcing our inclusion criteria, we had 14146 respondents for 2018 and 11383 for 2021. To give a preview of our data, @tbl-2018 and @tbl-2021 show the first five rows for each row for 2018 and 2021, respectively.
```{r}
#| label: fig-education-number-rows
#| fig-cap: Education
#| warning: false
#| message: false
#| echo: false
#| results: hide
print(nrow(filtered_df_2018))
print(nrow(filtered_df_2021))
```
```{r}
#| label: tbl-2018
#| tbl-cap: "Year 2018"
#| warning: false
#| message: false
#| echo: false
kable(filtered_df_2018[1:5,c("year","age","sex","marstat","racea","educ","inctot","k6sum")])
```
```{r}
#| label: tbl-2021
#| tbl-cap: "Year 2021"
#| warning: false
#| message: false
#| echo: false
kable(filtered_df_2021[1:5,c("year","age","sex","marstat","racea","educ","inctot","k6sum")])
```
## Measurement

@@

The MEPS survey collects participants' K6 score for nonspecific psychological distress. The K6 score is a summed scale value, also known as the Kessler 6 Scale, which considers six manifestations of nonspecific psychological distress. A score of 13 or higher indicates likely severe mental illness [@kessler_kessler_2012]. The variable called k6sum represents the score in our dataset [-@sec-appendix-data-details]. We constructed a new variable called "outcome," where 0 represents the score below 13 and 1 represents the score above 13. We considered zero as a person who is not experiencing psychological distress, denoted as "No," and one as a person experiencing severe psychological distress, denoted as "Yes." @fig-respondent-outcome represents the number of respondents we have in the dataset who were suffering from psychological distress. In 2018, 544 respondents were suffering severe psychological distress, where the number of total respondents was 14146. In 2021, 513 respondents were suffering from nonspecific psychological distress over the 30-day recall period, where the total number of respondents was 11383.

```{r}
#| label: fig-respondent-outcome
#| fig-cap: Number of Respondents Suffering Severe Psychological Distress
#| fig-subcap: 
#|  - "2018"
#|  - "2021"
#| layout-ncol: 2
#| echo: false
#| warning: false
max_count <- max(table(labeled_df_reduced_2018$outcome)) + 1000
labeled_df_reduced_2018 |> 
  ggplot(aes(x = outcome)) +
  geom_histogram(stat = "count", fill = "#013065", color = "white") +
  scale_y_continuous(limits = c(0, max_count)) +
  geom_text(
    stat = "count", 
    color = "black",
    aes(label = after_stat(count)), 
    vjust = -0.6,   
    size = 5        
  ) +
  labs(title = "", 
       x = "Outcome", 
       y = "Frequency") +
  theme_light()+
  theme(
    plot.title = element_text(size = 20, face = "bold"),   
    axis.title.x = element_text(size = 16),                
    axis.title.y = element_text(size = 16),                
    axis.text = element_text(size = 14)                    
  )

max_count <- max(table(labeled_df_reduced_2021$outcome)) + 1000
labeled_df_reduced_2021 |> 
  ggplot(aes(x = outcome)) +
  geom_histogram(stat = "count", fill = "#013065", color = "white") +
  scale_y_continuous(limits = c(0, max_count)) +
  geom_text(
    stat = "count", 
    color = "black",
    aes(label = after_stat(count)), 
    vjust = -0.3, 
    size = 5      
  ) +
  labs(title = "", 
       x = "Outcome", 
       y = "Frequency") +
  theme_light()+
  theme(
    plot.title = element_text(size = 20, face = "bold"),  
    axis.title.x = element_text(size = 16),               
    axis.title.y = element_text(size = 16),               
    axis.text = element_text(size = 14)                   
  )
```

@@

Our dataset has 7312 female respondents and 6834 male respondents for 2018. In 2021, the total number of respondents is lower compared to 2018. We have 6015 female respondents and 5368 male respondents for 2021 [@fig-sex].

```{r}
#| label: fig-sex
#| fig-cap: Distribution of Respondents by Sex (2018 and 2021)
#| layout-ncol: 1
#| echo: false
#| warning: false
# Join dataframe by sex
count_2018 <- labeled_df_reduced_2018 %>%
  group_by(sex) %>%
  summarise(Count = n()) %>%
  mutate(Year = "2018")

# Count gender distribution for 2021
count_2021 <- labeled_df_reduced_2021 %>%
  group_by(sex) %>%
  summarise(Count = n()) %>%
  mutate(Year = "2021")

combined <- bind_rows(count_2018, count_2021)
max_count <-  max(combined$Count) + 1000
ggplot(combined, aes(x = sex, y = Count, fill = Year)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.6) +
  scale_y_continuous(limits=c(0,max_count)) + 
    geom_text(aes(label = Count), 
            position = position_dodge(width = 0.6),
            vjust = -0.3,
            size = 5) +
  labs(
    title = "Gender Distribution: 2018 vs 2021",
    x = "Gender",
    y = "Count"
  ) +
  theme_light() +
  theme(
    plot.title = element_text(size = 18, face = "bold"),
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 16)
  )
```

@@

@fig-education-2018 and @fig-education-2021 depict the overall education status of our respondents. For both 2018 and 2022, a majority of respondents have at least a bachelor's, master's or PhD degree, followed by high school degree.
```{r}
#| label: fig-education-2018
#| fig-cap: Distribution of Respondents by Education (2018)
#| echo: false
#| warning: false

education_df_reduced_categories <- labeled_df_reduced_2018 |> 
  select_all() |>
  mutate(
    education=case_when(
      education=='Bachelor\'s'~'Bachelor\'s, Master\'s or PhD',
      education=='Master\'s'~'Bachelor\'s, Master\'s or PhD',
      education=='Doctoral'~'Bachelor\'s, Master\'s or PhD',
      TRUE~education
    )
  )
max_count <- max(table(education_df_reduced_categories$education)) + 1000
education_df_reduced_categories |> 
  ggplot(aes(x = education)) +
  geom_histogram(stat = "count", binwidth = 5, fill = "#013065", color = "white") +
  scale_y_continuous(limits = c(0, max_count)) +
  geom_text(
    stat = "count", 
    color = "black",
    aes(label = after_stat(count)), 
    vjust = -0.3,  # Adjust vertical position of labels
    size = 3       # Adjust size of labels
  ) +
  labs(title = "", 
       x = "Education", 
       y = "Frequency") +
  theme_light()+
  theme(
    plot.title = element_text(size = 15, face = "bold"),  
    axis.title.x = element_text(size = 10),               
    axis.title.y = element_text(size = 10),               
    axis.text = element_text(size = 10),             
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

@@

education_df_reduced_categories <- labeled_df_reduced_2021 |> 
  select_all() |>
  mutate(
    education=case_when(
      education=='Bachelor\'s'~'Bachelor\'s, Master\'s or PhD',
      education=='Master\'s'~'Bachelor\'s, Master\'s or PhD',
      education=='Doctoral'~'Bachelor\'s, Master\'s or PhD',
      TRUE~education
    )
  )

max_count <- max(table(education_df_reduced_categories$education)) + 1000
education_df_reduced_categories |> 
  ggplot(aes(x = education)) +
  geom_histogram(stat = "count", binwidth = 5, fill = "#013065", color = "white") +
  scale_y_continuous(limits = c(0, max_count)) +
  geom_text(
    stat = "count", 
    color = "black",
    aes(label = after_stat(count)), 
    vjust = -0.3,  
    size = 3       
  ) +
  labs(title = "", 
       x = "Education", 
       y = "Frequency") +
  theme_light()+
  theme(
    plot.title = element_text(size = 15, face = "bold"),
    axis.title.x = element_text(size = 10),        
    axis.title.y = element_text(size = 10),       
    axis.text = element_text(size = 10),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```
\newpage

@@

```{r}
#| label: fig-age
#| fig-cap: Distribution of Respondents by Age (2018, 2021)
#| echo: false
#| warning: false

all_age <-bind_rows(labeled_df_reduced_2021,labeled_df_reduced_2018)
max_count <- max(table(all_age$age)) + 1000
all_age |> 
  ggplot(aes(x = age)) +
  geom_histogram(fill = "#013065", color = "white") +
  scale_y_continuous(limits = c(0, max_count)) +
  labs(title = "", 
       x = "Age", 
       y = "Frequency") +
  theme_light()+
  theme(
    plot.title = element_text(size = 15, face = "bold"),
    axis.title.x = element_text(size = 10),
    axis.title.y = element_text(size = 10),
    axis.text = element_text(size = 10),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```
\newpage

@@

We use the logistic regression model to identify the association between suffering from psychological distress with sex and education. We set our model following the @tellingstories guidelines with the rstanarm package [@rstanarm_goodrich;@rstanarm_brilleman] in R [@citeR].

@@

We have found that psychological distress and education have a relation in terms of people who have higher education tend to feel lower psychological distress. Research indicates a strong connection between emotional well-being and education, as emotional learning is believed to be nurtured within the school environment and through the curriculum [@norwich_critical_2022;@billington_conditions_2022].

@@

We randomly selected 5000 samples for our model. We have not used any prediction to find out how good our model is for detecting psychological depression. Our next steps are carefully curating the dataset, introducing several data validation techniques, and increasing the sample set.
